---
title: "Data Appendix"
author: "Dägg and Janné, *Money, Happiness and the Midlife Crisis*"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: hide
    self_contained: yes
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r fileinfo, echo = FALSE}
### 
# Created by Johannes Karreth
# Last updated:  2018-05-24
# Written for R 3.5.0, Mac OS X 
### 
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# command_dir <- dirname(rstudioapi::getSourceEditorContext()$path)
# setwd(command_dir)
library("knitr")
library("tidyverse")
library("summarytools")
```

```{r dataprep, eval = FALSE, echo = FALSE}
# Note: These command files create the data used in this paper.
# eval is set to FALSE because these scripts need not be run each 
# time the paper is compiled.
source("../Command-Files/1-import-wdi.R")
source("../Command-Files/2-import-pew.R")
source("../Command-Files/3-processing.R")
```

```{r loaddata, echo = FALSE}
dc <- readRDS("../Analysis-Data/country-analysis.rds")
di <- readRDS("../Analysis-Data/individual-analysis.rds")
```

# Individual-level analysis data set: `individual-analysis.rds`

The individual-level data set contains observations representing `r nrow(di)` individuals from `r length(unique(di$CountryName))` countries. The distribution of observations across countries is as follows:

```{r freqtableCountryName, results = "asis"}
# table(di$CountryName)
freq(di$CountryName, report.nas = FALSE, style = "rmarkdown", omit.headings = TRUE)
```

The individual-level data set contains two individual-level variables: `SATIS` and `AGE`. Definitions, summary statistics and illustrations of the distributions of those two variables are shown below.

\clearpage

### Variable name: `SATIS`

- Source: Pew data set
- Values: Integers from 0 through 10
- Coding: Respondent's assessment of her/his overall satisfaction with life on a scale of 0 (least satisfied) to 10 (most satisfied)

```{r freqtableSatis, results = "asis"}
# table(di$satis)
freq(di$satis, report.nas = FALSE, style = "rmarkdown", omit.headings = TRUE)
```

```{r barSatis}
ggplot(data = di, aes(x = factor(satis))) + 
 geom_bar(aes(y = (..count..)/sum(..count..))) +
 scale_y_continuous() +
 xlab("Satisfaction (self-reported)") + 
 ylab("Share of respondents")
```

\clearpage

### Variable name: `AGE`

- Source: Pew data set
- Values: Positive integers
- Coding: Respondent's age in years at last birthday


```{r descrAge, results = "asis"}
# summary(di$age)
descr(di$age, stats = c("mean", "sd", "min", "Q1", "med", "Q3", "max"), transpose = TRUE, 
      omit.headings = TRUE, style = "rmarkdown")
```

```{r histAge}
ggplot(data = di, aes(x = age)) +
 geom_histogram(aes(y = (..count..)/sum(..count..)), binwidth = 1) +
 xlab("Age (in years)") + 
 ylab("Share of respondents")
```

\clearpage

# Country-level analysis data set: `country-analysis.rds`

The country-level data set contains one observation for each of the `r length(unique(di$CountryName))` countries included in the individual-level data set. The country-level data set contains three variables: `CM_SATIS`, `exp`, and `inc`. Definitions, summary statistics and illustrations of the distributions of those three variables are shown below.

### Variable name: `CM_SATIS`

- Source: Pew data set
- Values: Real numbers from 0 through 10
- Definition: Mean value of `SATIS` for all respondents from the country

```{r descrCmSatis, results = "asis"}
# summary(dc$cm_satis)
descr(dc$cm_satis, stats = c("mean", "sd", "min", "Q1", "med", "Q3", "max"), transpose = TRUE, 
      omit.headings = TRUE, style = "rmarkdown")
```

```{r histCmSatis}
ggplot(data = dc, aes(x = cm_satis)) +
  geom_histogram(binwidth = 1) +
  xlab("Country mean satisfaction") + 
  ylab("# of countries")
```

\clearpage

### Variable name: `exp`

- Source: WDI
- Values: Positive real numbers
- Definition: General government consumption expenditure for 2002, as a percentage of GDP


```{r descrExp, results = "asis"}
# summary(dc$exp)
descr(dc$exp, stats = c("mean", "sd", "min", "Q1", "med", "Q3", "max"), transpose = TRUE, 
      omit.headings = TRUE, style = "rmarkdown")
```

```{r histExp}
ggplot(data = dc, aes(x = exp)) +
  geom_histogram(binwidth = 4) +
  xlab("Government consumption, % of GDP") + 
  ylab("# of countries")
```

\clearpage

### Variable name: `inc`

- Source: WDI
- Values: Positive real numbers
- Definition: Income per capita for 2002, current US dollars


```{r descrInc, results = "asis"}
# summary(dc$inc)
descr(dc$inc, stats = c("mean", "sd", "min", "Q1", "med", "Q3", "max"), transpose = TRUE, 
      omit.headings = TRUE, style = "rmarkdown")
```

```{r histInc}
ggplot(data = dc, aes(x = inc)) +
  geom_histogram() +
  xlab("GDP per capita (current [2002] $ US)") + 
  ylab("# of countries")
```

```{r extractrcode, eval = FALSE, echo = FALSE}
# Note: this command extracts the R code from this .Rmd file and 
# creates a pure R script, then places it in the 
# Command-Files folder. Highlight and run the next two lines if you wish to 
# generate the R script.
purl(input = "3-data-appendix.Rmd", output = "4-data-appendix.R")
file.rename("4-data-appendix.R", "../Command-Files/4-data-appendix.R")
```
